---
title: "w3-fair-or-biased"
author: "CT Arrington"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Fair or unfair coins
There are two coins. One is fair with p=0.5 and the other is biased with p=0.7
Only one is being used.
The topic that we are estimating is "Is this coin fair or biased?"
Our prior belief is that the P(coin=biased)=0.6 and so P(coin=fair)=0.4


```{r, echo=TRUE}
bern <- function(p, y) {
  from_heads <- p^y
  from_tails <- (1-p)^(1-y)
  return (from_heads * from_tails)
} 

```

The likelihood of a heads is y=1 in the function $p^y(1-p)^{(1-y)}$
So the likelihood of a heads when coin=fair is ``r bern(0.5, 1)`` and when coin=biased it is
``r bern(0.7, 1)``
The likelihood of a tail when coin=fair is ``r bern(0.5, 0)`` and when coin=biased it is ``r bern(0.7, 0)`` 

The prior distribution of the coin is    
p(y) = $\sum_{coins}p(y|coin)p(coin)$ =    
p(y|coin=fair)p(coin=fair) + p(y|coin=biased)p(coin=biased)    

p(tails) = 0.5 * 0.4 + 0.3 * 0.6 = ``r 0.5 * 0.4 + 0.3 * 0.6``    
p(heads) = 0.5 * 0.4 + 0.7 * 0.6 = ``r 0.5 * 0.4 + 0.7 * 0.6``    


