---
title: "Equity in Admissions based on UCBAdmissions Dataset"
author: CT Arrington
output:
  pdf_document: default
  html_notebook: default
---

```{r, include=FALSE}
library(tidyverse)
library(kableExtra)
source("create_data.R")
source("calculate_thetas.R")
```

```{r, include=FALSE}
df = create_data()
original_df = as.tibble(UCBAdmissions)
original_thetas = calculate_thetas()
```

### Overall Admissions by Gender
The simplest way to think about equity in admissions is to consider the overall 
admission rates by gender.

As you can see below, men are accepted at a higher rate than women. So at first glance it seems clear that the admissons process
at UC Berkley was not equitable towards women in 1973.
```{r, echo=FALSE, out.width="50%"}
ggplot(original_df, aes(fill=Admit, y=n, x=Gender)) +
  geom_bar(position = "stack", stat = "identity")
```

### Overall Admissions by Department
The simple model of only looking at gender only makes sense if the admissions are very similar from department to department.
As you can see below, this is not at all the case. Some departments attract a lot more applicants than others
and some are much more competitive.
```{r, echo=FALSE, out.width="50%"}
ggplot(original_df, aes(fill=Admit, y=n, x=Dept)) +
  geom_bar(position = "stack", stat = "identity")
```

### Admissions by Gender for Each Department
Another way to look at equity is per department.

```{r, echo=FALSE, out.width="50%"}
ggplot(original_df, aes(fill=Admit, y=n, x=Gender)) +
  geom_bar(position = "stack", stat = "identity") +
  facet_wrap(vars(Dept))
```

### Admission Rates by Gender for Each Department
```{r, echo=FALSE, out.width="50%"}
ggplot(original_thetas, aes(fill=Gender, y=admission_rate, x=Gender)) +
  geom_bar(position = "stack", stat = "identity") +
  facet_wrap(vars(Dept))
```
### Next
```{r}
```